---
title: "Definitions"
author: "Julian Flowers"
date: "02/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rvest)
library(myScrapers)
library(readtext)
library(tidytext)

```

#### National Institute of Standards and Technology. NIST Big Data Interoperability Framework: Volume 1, Definitions (NIST Special Publication 1500-1). (2015). Available from: http://nvlpubs.nist.gov/nistpubs/SpecialPublications/NIST.SP.1500-1.pdf

Some big data and data science terms and definitions from the NIST Big Data definition manual.

```{r}

defs <- readtext("http://nvlpubs.nist.gov/nistpubs/SpecialPublications/NIST.SP.1500-1.pdf")

def_sents <- defs %>%
  unnest_tokens(sentences, text, token = "sentences") 
def_sents %>% 
  .[494:516, 2] %>%
  knitr::kable() %>%
  kableExtra::kable_styling()


```

* `r def_sents[494, 2]`
* `r def_sents[495, 2]`
* `r def_sents[496, 2]`
* `r def_sents[497, 2]`
* `r def_sents[498, 2]`
* `r def_sents[499, 2]`
* `r def_sents[500, 2]`
* `r def_sents[501, 2]`
* `r def_sents[502, 2]`
* `r def_sents[503, 2]`
* `r def_sents[505, 2]`
* `r def_sents[506, 2]`
* `r def_sents[507, 2]`
* `r def_sents[508, 2]`
* `r def_sents[509, 2]`
* `r def_sents[510, 2]`
* `r def_sents[511, 2]`
* `r def_sents[512, 2]`
* `r def_sents[513, 2]`
* `r def_sents[514, 2]`
* `r def_sents[515, 2]`
* `r def_sents[516, 2]`
* `r def_sents[517, 2]`








