---
title: "Rapid literature searching"
description: |
  Using R to rapidly search the literature 
author:
  - name: Julian Flowers 
    affiliation: Public Health England
    affiliation_url: https://www.gov.uk/phe
date: "`r Sys.Date()`"
output: 
   radix::radix_article:
     keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE )
```
 
Searching and critical appraising the literature is an essential skill for public health practitioners. Traditionally, this involves manual searching medical or bibliographic databases or sources of grey literature, identifying relevant abstracts or full text articles, reading and appraising retrieved papers and iterating through reference lists. If you are fortunate you might have access to a librarian or information specialist who can help with searching and retrieving publications. It can be a protracted manual process, and take many months.

With the development of APIs for bibliographic databases and R packages to read them it has become possible to automate searching and literature extraction, and with mainstreamed artificial intelligence tools for natural langauge processing (NLP) it has become possible to rapidly assess large numbers of papers.

## Example

For this exercise we'll need to download the following packages

* `tidyverse`
* `tidytext`
* `myScrapers`
* `quanteda`
* `Rcrawler`
* `rvest`


```{r}

library(myScrapers)

search <- "public health[MeSH] deep learning[tw]"
start <- 2000
end <- 2018
n <- 336

r_ph <- pubmedAbstractR(search = search, start = start, end = end, n = n)

r_ph

street_view_article <- readtext::readtext("https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5740675/pdf/pnas.201700035.pdf")

street_view_article$text

supp_text <- readtext::readtext("http://www.pnas.org/content/pnas/suppl/2017/11/27/1700035114.DCSupplemental/pnas.1700035114.sapp.pdf")
supp_text$text %>% head()
```


## Google searching

```{r}

doi <- googlesearchR(as.character(r_ph[51, "DOI"], n = 1))

doi[1]

get_page_text(as.character(doi[1]))




```

## Searching Pubmed

```{r}



```



### Searching Medline




